---
title: "Locosselli et al. PNAS modeling analysis"
author: "Melina Leite"
date: "`r format(Sys.time(), '%d de %B de %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: false
  pdf_document:
    highlight: tango
    toc: yes
---

```{r setup, echo=FALSE, warning=FALSE, message=F}
library(knitr);
#library(ape); library(phytools)


opts_chunk$set(fig.align = 'center', warning = F, message = F, error = F)
options(formatR.arrow = TRUE, width = 90)
```

```{r}
# libraries
library(MCMCglmm); library(coda)
library(parallel)
library(phytools); library(brranching); library(ape)
library(dplyr)
```

# Dataset

```{r}
df <- read.csv("data/dataset.csv")
str(df)
```

Columns information/metadata:

- zone: global climatic zone  
- ID: identity of the observation data  
- fam: Family name  
- taxagroup: taxonomical group, Conifer or Eudicot  
- species: Species name  
- longevity: recorde maximum longevity  
- growth: recorder growth  
- densityCHAVE: wood density  
- human.influ: Human influence index  
- tseas:  temperature seasonality
- moistwetq: water-balance soil moisture index of the wettest quarter of the year  
- low.w.rad: lowest weekly radiation  
- temp: mean annual temperature  
- moistdq:moisture index of the driest quarter   
- cloud: cloud cover  
- soil_class: soil classification
- alt.group: separation of the populations in "High altitude" and "Lowland"  
- phylo: Species names used to match the phylogenetic tree  

Standardizing numeric variables:
```{r}
data <- df %>% mutate_at(vars(8:15), scale)
```


## Phylogeny

The phylogenetic tree was constructed using the Phylomatic software (Webb et al. 2008) based on a revised vascular megatree (Gastauer & Meira-Neto 2016). We estimated the ancestral values of longevity and growth at internal nodes using maximum likelihood under the assumption of Brownian motion for trait evolution (Revel 2013). 
Auxiliary packages used: `phytools`, `brranching` and `ape`.

```{r}
filo <- read.tree("data/phylogenetic_tree.new")
```

Inverse phylogenetic matrix to be used in the model
```{r}
inv.phylo <- inverseA(filo, nodes = "TIPS", scale = FALSE)
```

# Model 1

Code of the bivariate phylogenetic linear mixed model with the packages `MCMCglmm` and `coda`.

The model takes long yours (or days) to run. The final version was stored to be loaded:
```{r}
load("model_output/modelResults.Rdata")
```

Prior specification for the random effects and residuals

```{r, eval=F}
prior <- list(R = list(V = diag(2), nu = 1.002),
               G = list(G1 = list(V = diag(2), nu = 1.002), # phylogeny
                        G2 = list(V = diag(2), nu = 1.002), # species
                        G3 = list(V = diag(2), nu = 1.002))) # soil_class
```

Model
```{r, eval = F}
mod <- MCMCglmm(cbind(log(longevity), log(growth)) ~ trait
        + trait:human.influ
        + trait:densityCHAVE 
        + trait:moistwetq + trait:temp + trait:moistdq
        + trait:tseas + trait:low.w.rad + trait:cloud - 1,
        random = ~ us(trait):phylo + us(trait):species + idh(trait):soil_class,
        rcov = ~ us(trait):units,
        family = c("gaussian", "gaussian"), 
        ginverse = list(phylo = inv.phylo$Ainv),
        prior = prior, data = data, pr = TRUE,
        nitt = 1000000, burnin = 10000, thin = 200)
```

Model summary
```{r}
summary(mod)
```

## Gelman-Rubin criteria of convergence (4 chains)

4 chains of the model previously ran to make the diagnose of the convergence of the chain:
```{r, echo=F}
load("model_output/modelResults_4chains.Rdata") 
mcov4 <- p1.4.sol
mcov4.var <- p1.4.var
```

```{r}
gelman.diag(mcov4[,1:18])
```

## Trace and density plots for fixed effects
```{r}
par(mfrow=c(4,2), mar=c(2, 1, 1, 1))
plot(mcov4[,1:18], ask=F, auto.layout=F)
```

## Trace and density plots for random effects
```{r}
par(mfrow=c(4,2), mar=c(2, 1, 1, 1))
plot(mcov4.var, ask=F, auto.layout=F)
```

# Model 2 lowland

Subsetting data:
```{r}
datalow <- data[data$alt.group == "Lowland" & df$temp >=20, ]
```


Prior specification for the random effects and residuals

```{r, eval=F}
prior <- list(R = list(V = diag(2), nu = 1.002),
               G = list(G1 = list(V = diag(2), nu = 1.002), # phylogeny
                        G2 = list(V = diag(2), nu = 1.002), # species
                        G3 = list(V = diag(2), nu = 1.002))) # soil_class
```

Model
```{r, eval=F}
mod.low20 <- MCMCglmm(cbind(log(longevity), log(growth)) ~ trait +
                      + trait:temp + trait:temp2 + trait:temp3 - 1,
                      random = ~ us(trait):phylo + us(trait):species + idh(trait):soil_class,
                      rcov = ~ us(trait):units,
                      family = c("gaussian", "gaussian"), ginverse = list(phylo = inv.phylo$Ainv),
                      prior = prior, data = datalow, pr = TRUE,
                      nitt = 100000, burnin = 10000, thin = 200)
```

Model's results previoulsy ran:
```{r}
load("model_output/modelResult_lowland.Rdata")
```

Model summary
```{r}
summary(mod.low20)
```

## Trace and density plots for fixed effects

```{r}
par(mfrow=c(4,2), mar=c(2, 1, 1, 1))
plot(mod.low20$Sol[,1:8], ask=F, auto.layout=F)
```

## Trace and density plots for random effects

```{r}
par(mfrow=c(4,2), mar=c(2, 1, 1, 1))
plot(mod.low20$VCV, ask=F, auto.layout=F)
```


# Session Information
```{r}
sessionInfo()
```

# References

Webb C.O., D. D. Akerly, S. W. Kembel, Phylocom: software for the analysis of phylogenetic community structure and character evolution. Bioinformatics, 24, 2098-2100. http://dx.doi.org/10.1093/bioinformatics/btn358. PMid:18678590 (2008).

Gastauer,M., J. A. A. Meira-Neto, An enhanced calibration of a recently released megatree for the analysis of phylogenetic diversity. Brazilian Journal of Biology, 76(3), 619-628 (2016).

Revell, L.J. Two new graphical methods for mapping trait evolution on phylogenies. Methods in Ecology and Evolution 4, 754–759 (2013).

